Corpus Experiments for Word Embeddings

### 2015-08-18 article/main.tex
spell checked

% compile using latex (xelatex seems to crop the figure strangley)
\documentclass{article} % For LaTeX2e
\usepackage[dvips]{graphicx}
\usepackage{nips14submit_e,times}
\usepackage{hyperref}
%\usepackage{wrapfig}
\usepackage{rotating}
\usepackage[outdir=./]{epstopdf}

@@

\begin{abstract}
	We introduce an experimental approach to studying the properties of word embeddings.
	Controlled experiments, acheived by modification of the training corpus, permit the demonstration of direct, causal relationships between word properties and word vector direction and length.
	This approach is demonstrated in the case of the word2vec CBOW model with experiments that independently vary word frequency and word co-occurrence noise.
	It is shown in this case that word vector length depends directly and essentially linearly on both word frequency and the level of noise in the co-occurrence distribution of the word.
	In both cases, the coefficient of linearity depends upon the word.
\end{abstract}

@@

In this short note, we perform experiments that illustrate the effect of these two determining pieces of information (the co-occurrence distribution and the word frequency) as they are varied independently of one another.
This is acheived through the introduction of new tokens into the training corpus with varying frequencies and varying levels of noise in their co-occurrence distributions.
The frequency and co-occurrence distributions of these introduced tokens are modelled on existing words in the corpus.

@@

This short note is structured as follows.
Section \ref{related-work} draws connections to related work, while Section \ref{corpus-and-model} describes the corpus and model used to illustrate the experiments.
Section \ref{WFVE} describes controlled experiment for varying word frequency while holding the co-occurrence distribution fixed.
Section \ref{CNVE}, in a complementary fashion, describes a controlled experiment for varying the level of noise in the cooccurrence distribution of a word, while holding the word frequency fixed.
Section \ref{future-directions} considers further questions and possible future directions.
\end{section}

@@

For simplicity, only the word2vec CBOW word embedding with a single set of hyperparameters is considered.
Specifically, a $100$-dimensional CBOW model is trained using negative sampling with 5 negative samples, a window size of $10$, a minimum word occurrence of $128$, and $10$ passes through the corpus.
Subsampling was not used so that the influence of word frequency could be more clearly discerned.
Identical experimental results were obtained using hierarchical softmax, but these are omitted for succinctness.
The relatively high minimum word cut-off is chosen to ensure that word vectors receive a sufficient number of gradient updates to be meaningful.
\textcolor{red}{This cut-off results in a vocabulary of $62025$ words (only unigrams were considered).}

@@

\begin{figure}
	\input{word-frequency-experiment-text-cat.tex}
	\caption{A text modified for the word frequency experiment as per subsection \ref{WFVEexisting}, where the
	word \word{cat} was chosen, $\lambda=0.5$ and $n=20$.}
	\label{fig:word-frequency-experiment-text-cat}
\end{figure}

@@

\begin{figure}\label{fig:word-frequency-experiment-text-void}
	\input{word-frequency-experiment-text-void.tex}
	\caption{A text modified for the word frequency experiment as per subsection \ref{WFVEmeaningless}, where $\lambda=0.5$ and $n=20$. For illustrative purposes, the token \word{VOID} was here interspersed with a global frequency of $0.05$.}
\end{figure}
\end{subsection}

\subsection{Experimental results}
Word frequency has no effect on word vector direction.
Figure \ref{word-frequency-experiment-heatmap} shows the cosine similarity of the pairs of word vectors for the tokens introduced into the corpus.
The cosine similarity measures the extent to which two vectors have the same direction, taking a maximum value of $1$ and a minimum value of $0$.
Notice that the word vectors associated to tokens derived from meaningful tokens have the same direction.
The same is true of the word vectors associated to the \word{VOID\_i} for $i$ \textcolor{red}{below $7$ or $8$}, but that the word vectors \word{VOID\_i} gradually diverge directionally for higher values of $i$.
This demonstrates the difficulty of learning the co-occurrence distribution of $\word{VOID}$, the global frequency distribution.
Because of its noisiness, a higher number of samples (occurrences) is required to learn this co-occurrence distribution.
The same holds, but to a lesser extent, for the co-occurrence distribution of the stopword \word{the}.

It remains, therefore, to consider the effect of frequency variation on word vector length.
Figure \ref{fig:frequency-norm-graph} illustrates the global relationship between vector length and word frequency.
Figure \ref{fig:word-frequency-experiment-graph}, on the other hand, shows this relationship for individual words, both for the word vectors, ``syn0'' and for the discarded second layer vectors ``syn1neg'' (included for completeness).
Each line corresponds to a single word, and the points on each line indicate the frequency and vector length for the tokens derived from the corresponding word.
\textcolor{red}{For example, the five points on the line corresponding to \word{admiral} are labelled, from right to left, by the tokens \word{ADMIRAL\_1}, \word{ADMIRAL\_2}, \dots, \word{ADMIRAL\_5}.}
The number of points on the line is determined by the frequency of the original word.
\textcolor{red}{For example, the frequency of the \word{admiral} can be halved at most $5$ times and remain about the minimum frequency cut-off.}

@@

\begin{section}{Co-occurrence noise variation experiment}\label{CNVE}
Here, and throughout, the noise distribution is taken to be the global word frequency distribution.
Thus noise can be added to the co-occurrence distribution of a word by inspersing occurrences of that word uniformly at randomly throughout the corpus.
A small set of words is chosen from the unmodified corpus in the same manner as in section \ref{WFVE}.

@@

The proportion of the occurrences of \word{CAT\_i} that arose by replacing \word{cat} is $P_{5/6}(i)$, and the proportion arising from the random intersperal of this token throughout the corpus is $r_i := (1 - P_{5/6}(i))$.
The particular value of $\lambda$ (large compared to that of section \ref{WFVE}) was chosen so that the proportion of noise $r_i$ did not increase too quickly with $i$.
These proportions are the sampling points on our axis of variation, and this value of $\lambda$ (together with the choise of $n$) ensures a reasonable spread.
Note that other parameter values (or indeed other distributions) could equally have been used.

@@

\begin{figure}
	\input{cooccurrence-noise-experiment.tex}
	\caption{A text modified for the co-occurrence noise experiment, where the word \word{cat} was chosen, $\lambda = 5/6$ and $n=3$.}
\label{fig:co-occurrence-noise-experiment-text}
\end{figure}

>>>>

% compile using latex (xelatex seems to crop the figure strangely)
\documentclass{article} % For LaTeX2e
\usepackage[dvips]{graphicx}
\usepackage{nips14submit_e,times}
\usepackage{hyperref}
\usepackage{mdframed}
%\usepackage{wrapfig}
\usepackage{rotating}
\usepackage[outdir=./]{epstopdf}

@@

\begin{abstract}
	We introduce an experimental approach to studying the properties of word embeddings.
	Controlled experiments, achieved by modification of the training corpus, permit the demonstration of direct, causal relationships between word properties and word vector direction and length.
	This approach is demonstrated in the case of the word2vec CBOW model with experiments that independently vary word frequency and word co-occurrence noise.
	It is shown in this case that word vector length depends directly and essentially linearly on both word frequency and the level of noise in the co-occurrence distribution of the word.
	In both cases, the coefficient of linearity depends upon the word.
\end{abstract}

@@

In this short note, we perform experiments that illustrate the effect of these two determining pieces of information (the co-occurrence distribution and the word frequency) as they are varied independently of one another.
This is achieved through the introduction of new tokens into the training corpus with varying frequencies and varying levels of noise in their co-occurrence distributions.
The frequency and co-occurrence distributions of these introduced tokens are modeled on existing words in the corpus.

@@

This short note is structured as follows.
Section \ref{related-work} draws connections to related work, while Section \ref{corpus-and-model} describes the corpus and model used to illustrate the experiments.
Section \ref{WFVE} describes controlled experiment for varying word frequency while holding the co-occurrence distribution fixed.
Section \ref{CNVE}, in a complementary fashion, describes a controlled experiment for varying the level of noise in the co-occurrence distribution of a word, while holding the word frequency fixed.
Section \ref{future-directions} considers further questions and possible future directions.
\end{section}

@@

For simplicity, only the word2vec CBOW word embedding with a single set of hyperparameters is considered.
Specifically, a $100$-dimensional CBOW model is trained using negative sampling with 5 negative samples, a window size of $10$, a minimum word occurrence of $128$, and $10$ passes through the corpus.
Sub-sampling was not used so that the influence of word frequency could be more clearly discerned.
Identical experimental results were obtained using hierarchical softmax, but these are omitted for succinctness.
The relatively high minimum word cut-off is chosen to ensure that word vectors receive a sufficient number of gradient updates to be meaningful.
\textcolor{red}{This cut-off results in a vocabulary of $62025$ words (only unigrams were considered).}

@@

\begin{figure}\label{fig:word-frequency-experiment-text-cat}
	\begin{mdframed}
	\input{word-frequency-experiment-text-cat.tex}
	\end{mdframed}
	\caption{A text modified for the word frequency experiment as per subsection \ref{WFVEexisting}, where the
	word \word{cat} was chosen, $\lambda=0.5$ and $n=20$.}
\end{figure}

@@

\begin{figure}\label{fig:word-frequency-experiment-text-void}
	\begin{mdframed}
	\input{word-frequency-experiment-text-void.tex}
	\end{mdframed}
	\caption{A text modified for the word frequency experiment as per subsection \ref{WFVEmeaningless}, where $\lambda=0.5$ and $n=20$. For illustrative purposes, the token \word{VOID} was here interspersed with a global frequency of $0.05$.}
\end{figure}
\end{subsection}

\subsection{Experimental results}
Word frequency has no effect on word vector direction.
Figure \ref{word-frequency-experiment-heatmap} shows the cosine similarity of the pairs of word vectors for the tokens introduced into the corpus.
The cosine similarity measures the extent to which two vectors have the same direction, taking a maximum value of $1$ and a minimum value of $-1$.
Notice that the word vectors associated to tokens derived from meaningful tokens have the same direction.
The same is true of the word vectors associated to the \word{VOID\_i} for $i$ \textcolor{red}{below $7$ or $8$}, but that the word vectors \word{VOID\_i} gradually diverge directionally for higher values of $i$.
This demonstrates the difficulty of learning the co-occurrence distribution of $\word{VOID}$, the global frequency distribution.
Because of its noisiness, a higher number of samples (occurrences) is required to learn this co-occurrence distribution.
The same holds, but to a lesser extent, for the co-occurrence distribution of the stopword \word{the}.

It remains, therefore, to consider the effect of frequency variation on word vector length.
Figure \ref{fig:frequency-norm-graph} illustrates the global relationship between vector length and word frequency.
Figure \ref{fig:word-frequency-experiment-graph}, on the other hand, shows this relationship for individual words, both for the word vectors, ``syn0'' and for the discarded second layer vectors ``syn1neg'' (included for completeness).
Each line corresponds to a single word, and the points on each line indicate the frequency and vector length for the tokens derived from the corresponding word.
\textcolor{red}{For example, the five points on the line corresponding to \word{admiral} are labeled, from right to left, by the tokens \word{ADMIRAL\_1}, \word{ADMIRAL\_2}, \dots, \word{ADMIRAL\_5}.}
The number of points on the line is determined by the frequency of the original word.
\textcolor{red}{For example, the frequency of the \word{admiral} can be halved at most $5$ times and remain about the minimum frequency cut-off.}

@@

\begin{section}{Co-occurrence noise variation experiment}\label{CNVE}
Here, and throughout, the noise distribution is taken to be the global word frequency distribution.
Thus noise can be added to the co-occurrence distribution of a word by interspersing occurrences of that word uniformly at randomly throughout the corpus.
A small set of words is chosen from the unmodified corpus in the same manner as in section \ref{WFVE}.

@@

The proportion of the occurrences of \word{CAT\_i} that arose by replacing \word{cat} is $P_{5/6, 8}(i)$, and the proportion arising from the random interspersal of this token throughout the corpus is $r_i := (1 - P_{5/6, 8}(i))$.
The particular value of $\lambda$ (large compared to that of section \ref{WFVE}) was chosen so that the proportion of noise $r_i$ did not increase too quickly with $i$.
These proportions are the sampling points on our axis of variation, and this value of $\lambda$ (together with the choice of $n$) ensures a reasonable spread.
Note that other parameter values (or indeed other distributions) could equally have been used.

@@

\begin{figure}
	\begin{mdframed}
	\input{cooccurrence-noise-experiment.tex}
	\end{mdframed}
	\caption{A text modified for the co-occurrence noise experiment, where the word \word{cat} was chosen, $\lambda = 5/6$ and $n=3$.}
\label{fig:co-occurrence-noise-experiment-text}
\end{figure}
